Staged Abstract Interpreters

### 2019-08-11 paper/10_intro.tex
update numbers
revise intro

Statically analyzing semantic properties of a program is a
widely-known undecidable problem. Abstract interpretation as a
lattice-based approach to sound static analyses was proposed
by \citet{DBLP:conf/popl/CousotC77}. With the Galois connections, one
can approximate to the program runtime behaviors by computing the
fixed points on the abstract domain. Despite the tremendous
theoretical development of abstract interpretation over the years, on
the side of pragmatics, constructing artifacts and analyzers that
perform sound abstract interpretation for modern, high-level and
expressive languages was considered abstruse and complicated for a
long time.

In recent years, we observe a rich progress on the methodologies for
constructing abstract interpreters from systematic principles, instead
of ad-hoc engineering. A notable one is the Abstracting Abstract
Machine (AAM) methodology \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, which uncovers an approach to derive sound
abstract interpreters from their concrete counterparts, i.e., abstract
machines, where the soundness can be easily established by examining
the transformation of semantic artifacts. For example, a CEK machine
\cite{DBLP:conf/popl/FelleisenF87} for concrete execution can be
readily refactored to an effective $0$-CFA control-flow analysis
\cite{Shivers:1988:CFA:53990.54007,
  Midtgaard:2012:CAF:2187671.2187672} by first tweaking the
environment dereference as a nondeterministic choice, then allocating
continuations in the environment, and constraining the addresses space
to be finite. This systematic abstraction approach can be tailored to
various language features (such as states, first-class controls,
exceptions and concurrency) and different sensitivity analyses
\cite{DBLP:conf/icfp/Gilray0M16, DBLP:conf/popl/GilrayL0MH16,
  Darais:2015:GTM:2814270.2814308}, and has been applied to multiple
small-step abstract machines \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10, Sergey:2013:MAI:2491956.2491979} and
big-step definitional interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

Based on the idea of AAM, and more pragmatically, as an implementation
strategy, several purely functional programming approaches to build
abstract interpreters were emerged, for examples, using monads or
arrows. The pure approach provides much benefits: 1) the abstract
interpretation artifacts are compositional and can be constructed
modularly, e.g., by using monad transformers. 2) The soundness of
analysis can be proved more easily, by either mechanization
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proof
\cite{Keidel:2018:CSP:3243631.3236767}.  3) The correctness of
implementations can be reasoned by the programmers more confidently,
e.g., by leveraging equational reasoning.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties with these high-level programming approaches.
First, similar to concrete interpreters, the abstract interpreter analyzes the
program by traversing the abstract syntax tree, which poses the interpretive
overhead, e.g., pattern matching on the ASTs and recursive calls on the sub
expressions. Those kind of overhead can be negligible if the abstract
interpreter only runs on the program once, but also can be accumulated
significantly if it runs repeatedly, for example, on libraries. Second, the
abstract interpreter written in pure languages usually extensively uses effect
systems to model the behaviors of abstract interpretation, such as
nondeterminism. For example, \citet{DBLP:journals/pacmpl/DaraisLNH17} and
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract interpreters;
the very recent \citet{Githubsemantic}'s \textsc{Semantic} framework
uses extensible effects. With certain merits and elegance of the implementation,
the effect systems may also introduce additional performance overhead.

In this paper, we propose an abstraction-without-regret approach to
eliminate those performance penalties for abstract interpreters, while
still keeping these benefits from purely functional programming.  In
short, our approach borrows ideas from program specialization and
embedded DSLs, and applies them to this particular kind of
meta-programs, abstract interpreters.  1) Using multi-stage
programming , we can specialize the
abstract interpreter with respect to an input program and then
generate efficient low-level code that does the actual analysis. The
result of specialization is reusable, and the effect layers have been
eliminated in the generated code. 2) We use a tagless-final approach
to embed the target language interpreter in a high-level host language,
which defines a generic interpreter and allows user to
implement different semantics modularly, including the abstract
semantics and staging, without changing the generic one.
Together with the multi-stage programming and type-level
annotations, this design allows user to derive staged abstract
interpreters without intrusive changes to the unstaged one, thus the soundness is preserved. 
Therefore, our approach is no regret in the sense of both performance and least
engineering efforts to achieve such performance. We elaborate these
two main ideas in detail as follow.

@@

The idea of specializing interpreters can be traced to Futamura
projections \cite{Futamura1999, futamura1971partial}.
The first Futamura projection specifically shows that
specializing an interpreter w.r.t. the input program yields an
equivalent executable.  For instance, we have an interpreter @eval@ of
some language:
\begin{lstlisting}
  def eval(e: Expr)(arg: Input): Value
\end{lstlisting}
Given a program $e_0$ of type @Expr@, by applying the first Futamura
projection, we can obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ : @Input@ $\to$ @Value@, which is the so-called
\textit{equivalent executable}. By definition of the interpreter,
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first
proposed approach to realize Futamura projections: it discovers the
binding-time (static or dynamic) of variables, evaluates the static
part, and generates a residualized program that solely relies on the dynamic
part. However, it is hard to precisely analyze binding-times given an
arbitrary program. As an alternative approach to specialization and
partial evaluation, multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97}
lets the programmer to explicitly control and annotate the
binding-times in the program, and the MSP system will specialize the
program. The staging annotations can be either syntactic (e.g.,
MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} in Scala).

Our proposed framework adopts the type-based multi-stage programming
and implements the first Futamura projection of a big-step abstract
interpreter for a stateful, higher-order core language.  We use
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s abstract definitional
interpreter that uses monad transformers as the unstaged baseline. By
deriving staged monads that can be used to do code generation, we may
obtain a staged abstract interpreter that fulfills the specialization.
After staging, the generated code is specialized to the input program,
and all monadic operations are all inlined and compiled down to low-level
Scala code.

@@

\paragraph{Applications and Evaluations}
We evaluate the idea of staging an abstract interpreter through
multiple case studies and empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, and show that we can achieve
the same optimization by utilizing type-based stage annotations;
meanwhile, the analyzer program does not need any changes, thereby
requires little engineering efforts.
2) We extend the basic staged abstract interpreter to diverse flow
analysis, including a store-widened analysis, a context-sensitive
analysis and abstract garbage collection.
3) We notice that staging an abstract interpreter enables modularly
compiling an analysis to programs. The compiled analysis is reusable,
therefore it provides a mechanized way to create library summaries by
reusing a whole-program analyzer.
4) We also empirically evaluate the performance improved by staging,
showing an average 11x speed-up on flow-analysis.

>>>>

Statically analyzing semantic properties of a program is a
widely-known undecidable problem. Abstract interpretation as a
lattice-based approach to sound static analyses was proposed
by \citet{DBLP:conf/popl/CousotC77}. Equipped with Galois connections,
the analyzer can obtain the program runtime behaviors approximately
by computing the fixed points on the abstract domain. Despite the tremendous
theoretical development of abstract interpretation over the years,
constructing artifacts and analyzers that perform sound abstract
interpretation for modern and expressive languages was considered
abstruse and complicated for a long time.

In recent years, we observe a rich progress on the methodologies for
constructing abstract interpreters from systematic principles, instead
of ad-hoc engineering. A notable one is the Abstracting Abstract
Machine (AAM) methodology \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}. It uncovers an approach to derive sound
abstract interpreters from their concrete counterparts, for example,
abstract machines, where the soundness can be easily established by examining
the transformation of semantic artifacts. For example, the CEK machine
\cite{DBLP:conf/popl/FelleisenF87} for concrete execution can be
readily refactored to an effective $0$-CFA control-flow analysis
\cite{Shivers:1988:CFA:53990.54007, Midtgaard:2012:CAF:2187671.2187672}:
first we tweak the environment dereference as a nondeterministic choice,
such that the environment may contain mutiple possible values for a varaiable,
then allocating continuations in the environment to exploit return-flow information,
and also constrain the addresses space to be finite.
This systematic abstraction approach can be tailored to
different language features (such as states, first-class controls,
exceptions and concurrency) and sensitivity analyses
\cite{DBLP:conf/icfp/Gilray0M16, DBLP:conf/popl/GilrayL0MH16,
  Darais:2015:GTM:2814270.2814308}. It also has been applied to various
small-step abstract machines \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10, Sergey:2013:MAI:2491956.2491979} and
big-step definitional interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

Based on the idea of abstracting abstract machines, more pragmatically,
several implementation strategies utilizing purely functional programming
to build abstract interpreters were emerged. Such techniques include monads,
monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979},
arrows \cite{Keidel:2018:CSP:3243631.3236767}, extensible effects \cite{Githubsemantic} and etc.
The pure approaches provide certain benefits. The abstract interpretation
artifacts can be built compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proved with less efforts,
either by mechanized \cite{Darais:2016:CGC:2951913.2951934} or paper-based proof
\cite{Keidel:2018:CSP:3243631.3236767}. Also, referential transparent and
equational reasoning allow programmers to reason the correctness of their
implementations more confidently.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties with these high-level implementation approaches.
First, similar to concrete interpreters, the abstract interpreter analyzes the
program by traversing the abstract syntax tree, which poses an interpretive
overhead, such as the pattern matching on the ASTs and recursive calls on the sub
expressions. Those kind of overhead can be negligible if the abstract
interpreter only runs on the program for one time, but also can be accumulated
significantly if it runs repeatedly, for example, on libraries.
Second, the abstract interpreter written in pure languages usually extensively
uses effect systems to implement the semantics of abstract interpretation.
For example, the abstract interpreter that returns all possible runtime values
is a form of nondeterminism, where nondeterminism monads can help.
Although such pure approaches have its own merits and elegance,
compared with imperative stateful implementations, they are significantly slower.

In this paper, we propose an abstraction-without-regret approach to
eliminate those performance penalties for abstract interpreters, meanwhile
still keeping the benefits come from purely functional programming.
In short, our approach applies ideas and techniques from program specialization
and embedded DSLs to abstract interpreters. 1) Using multi-stage
programming , we can specialize the abstract interpreter with respect to an
input program and then generate efficient low-level code that does the actual analysis.
The result of specialization is reusable, and the effect layers have been
eliminated in the generated code. 2) Inspired by the tagless-final interpreters,
we define a generic interpreter that abstracts over binding-time and
different semantics, which allows user to implement different semantics
modularly, including the staged abstract interpretation semantics.
Together with the multi-stage programming and type-level annotations of binding-times,
we can derive the staged abstract interpreters without intrusive changes to
the corresponding unstaged one, thus the soundness is untouched.
In this sense, our approach is no regret of both performance and engineering
effort. We elaborate these two main ideas in detail as follow.

@@

The idea of specializing interpreters can be traced to Futamura
projections \cite{Futamura1999, futamura1971partial}.
The first Futamura projection specifically shows that
specializing an interpreter w.r.t. the input program yields an
equivalent executable. For instance, @eval@ is an interpreter for
some language:
\begin{lstlisting}
  def eval(e: Expr)(arg: Input): Value
\end{lstlisting}
Given a program $e_0$ of type @Expr@, by applying the specialization,
we can obtain a specialized interpreter $\texttt{eval}_{\texttt{e0}}$ :
@Input@ $\to$ @Value@, which is the so-called \textit{equivalent executable}.
By definition of the interpreter, they produce the same result when applied
with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first
proposed approach to realize Futamura projections: it first discovers
the binding-times (\textit{static} or \textit{dynamic}) of variables,
then evaluates the static part of the program, and generates a residualized
program that solely relies on the dynamic part. However, it is hard to
precisely analyze binding-times given an arbitrary program.
As an alternative approach to specialization, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97}
lets the programmer to explicitly control and annotate the
binding-times in the program, then the MSP system will check whether these annotations
are consistent and specialize the program using that information.
The staging annotations can be either syntactic (e.g., quote and quasiquote
in MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular Staging
framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

Our proposed framework adopts the type-based multi-stage programming from
the Lightweight Modular Staging framework and implements the Futamura
projection of a big-step abstract interpreter for a small stateful,
higher-order language.
We use \citet{DBLP:journals/pacmpl/DaraisLNH17}'s abstract definitional
interpreter that uses monad transformers as the unstaged baseline. By
deriving staged monads that can be used to do code generation, we may
obtain a staged abstract interpreter that fulfills the specialization.
After staging, the generated code is specialized to the input program,
and all monadic operations are all inlined and compiled down to low-level
Scala code.

@@

\paragraph{Applications and Evaluations}
We evaluate the idea of staging an abstract interpreter through
multiple case studies and empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, and show that we can achieve
the same optimization by utilizing type-based stage annotations;
meanwhile, the analyzer program does not need any changes, thereby
requires little engineering efforts.
2) We extend the basic staged abstract interpreter to diverse flow
analysis, including a store-widened analysis, a context-sensitive
analysis and abstract garbage collection.
3) We notice that staging an abstract interpreter enables modularly
compiling an analysis to programs. The compiled analysis is reusable,
therefore it provides a mechanized way to create library summaries by
reusing a whole-program analyzer.
4) We also empirically evaluate the performance improved by staging,
showing an order of magnitude speed-up on flow-analysis.

### 2019-08-11 paper/abstract.tex
revise intro

It is well known that a staged interpreter is a compiler: specializing the
interpreter to a given program produces an equivalent program that runs faster,
which is known as the first Futamura projection. It is even more widely known that an
abstract interpreter is a program analyzer: tweaking the interpreter to run on a
domain of abstract values produces a sound static analysis. What happens when we
combine these two ideas, and apply staging to an \emph{abstract} interpreter?

In this paper, we present a unifying framework that naturally extends the first
Futamura projection of concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstraction and
monadic abstraction. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in four modes: unstaged concrete
interpreter, staged concrete interpreter, unstaged abstract interpreter, or
staged abstract interpreter.

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the application of various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of 11x with staging for control-flow analysis.

>>>>

It is well known that a staged interpreter is a compiler: specializing the
interpreter to a given program produces an equivalent executable that runs faster.
This connection is known as the first Futamura projection.
It is even more widely known that an abstract interpreter is a program analyzer:
tweaking the interpreter to run on an abstract domain produces a sound static
analysis. What happens when we combine these two ideas, and apply staging to
an \emph{abstract} interpreter?

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstraction and
monadic abstraction. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in four modes: unstaged concrete
interpreter, staged concrete interpreter, unstaged abstract interpreter, or
staged abstract interpreter.
As an example of \emph{abstraction without regret}, the overhead of these
abstraction layers is eliminated in the generated code after staging.
We show that staging abstract interpreters is practical and useful to
optimize static analysis while requiring less engineering efforts and not
compromising soundness. We conduct three case studies, including a comparison
with \citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
applications on various control-flow analyses, and a demonstration shows that
it can be used for modular analysis.
We also empirically evaluate the running time improved by staging.
The experiment shows an order of magnitude speedup with staging for
control-flow analyses.
